Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-08-09 paper/letter.tex
update body

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We explain these features by two different models: an
homogeneous sphere undergoing gas bulk rotation and an interstellar
medium composed by clumps with random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming these physical conditions.
We  find that both models independently require high velocities that
translate into a dynamical mass at least $10$ times larger than the
neutral mass hydrogen in the galaxy. 
We argue that a possibility to explain the high
dynamical mass is the presence of a super-massive black hole. 
Ionization emission lines present in this galaxy support the idea that
\tol\ might have harbored an an Active Galactic Nucleus  
The implications of our findings for the study of LAEs, including the
first generation of galaxies are manifold. 
It demonstrates the importance of considering rotation and multiphase
physics under the possible conditions shaping the \lya\ spectra of the
first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of active supermassive black
holes.  
}

@@

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.   
Recent advances in these computational models have explored the
effects that bulk rotation and a multiphase state in the interstellar
medium should have on on the Lyman-$\alpha$ line.

The most important consequence of bulk rotation is that, even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is close to the half-line width of the
static line the line becomes single peaked and symmetric, a unique
feature that other theoretical models find difficult to reproduce
without introducing more complexity into the gas distribution.

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ had evaded a physical
interpretation so far, because bulk rotation and a multiphase ISM can
independently explain the \lya\ line in \tol.

Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continuous line) and a clumpy ISM
(dashed line).

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{-60}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $3.0^{+1.5}_{-1.8}\times
10^9$ M$_{\odot}$.

Radio surveys of the 21cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$ \cite{pustilnikmartin07}. 
In the case of the rotation model, this information help us to constrain the
diameter of the HI region where the \lya\ emission and transfer takes
place. 
We find this size to be in the range $0.11 < D/\mathrm{kpc}<0.34$, one
order of magnitude smaller than \tol's size in the optical.   
That size and the rotational velocity of $300$\kms put a constraint on
the dynamical mass of $4.5^{+2.1}_{-2.4} \times 10^{9}$M$_{\odot}$.

This is evidence for a dynamical mass at least 11 (multiphase case) to
17 (rotation case) times higher than the mass in neutral hydrogen.  
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars has been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

@@

From the multiphase model we obtained $v=71_{-25}^{+16}$ \kms
and $r=5$ kpc. 
This corresponds to a dynamical mass of $M_{\rm
  dyn}=5.9^{+3.1}_{-3.4}\times 10^{9}$ $M_{\odot}$ which is at least
$20$ times the HI mass estimated from observations.

In the rotation model the size of the spherical region, $r_{s}$, can
be infered from the relationship $\tau = \sigma_0 n r_s$, where
$\sigma_0=5.898\times 10^{-14}$cm$^{-2}$ is the Lyman$\alpha$
cross section at the  line's center, $n$ is the number density. 
This expresion can be rewritten as
\begin{equation}
r_{s} = 0.055 \left(\frac{\tau}{10^7}\right) \left(
\frac{\mathrm{atoms\ cm^{-3}}}{n}\right) \mathrm{kpc}.
\end{equation}
%
Using this result and assuming a bound for the number density of
$10^{-2}<n/\mathrm{atoms\ cm^{-3}}<1$ we have $5.5\times 10^{-2}
< r_s/\mathrm{kpc}<5.5$.

@@

Using those size bounds for the spherical model we find a constrain for
the dynamical mass of $5.45\times 10^{8}M_{\odot} < M_{\rm dyn} <
5.45\times 10^{10} M_{\odot}$.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MG$^3$, MCRG$^1$, JNGC$^2$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
In this letter we explain these features by two different models: an
homogeneous gaseous sphere undergoing bulk rotation and an interstellar
medium composed by clumps with random motions.
It is the first time that an observed \lya\ spectrum can be explained
assuming either of these physical conditions.
We find that both models independently require high velocities that
translate into a dynamical mass at least $20$ times larger than the
neutral mass hydrogen in the galaxy. 
We argue that a plaussible explanation for the high dynamical mass is
the presence of a super-massive black hole.  
This work demonstrates the importance of considering rotation and
multiphase physics under the possible conditions shaping the
\lya\ spectra of the first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black
holes.  
}

@@

However, the physical interpretation of Ly-$\alpha$ observations is
not straightforward. 
This is due to the resonant nature of the Ly-$\alpha$ line. 
A Ly-$\alpha$ photon follows a diffusion-like process before escaping
the galaxy or being absorbed by dust. 
The resulting line profile becomes sensitive to the dynamical, chemical
and thermal conditions in the interstellar medium. 
There are very few analytically tools available to interpret the
Ly-$\alpha$ line.
They are applicable only in very few cases of highly symmetrical
conditions, which are hardly met in real astrophysical systems.
For these reasons the interpretation of Ly-$\alpha$ observations
require state-of-the-art Monte Carlo radiative transfer simulations.   
Recent advances in these computational models have explored the
effects that bulk rotation and a multiphase state in the interstellar
medium should have on the Lyman-$\alpha$ line.

The most important consequence of bulk rotation is that, even for a
spherical gas distribution, the line morphology now depends on the
viewing angle respect to the rotation axis.  
For a line of sight perpendicular to the rotation axis the intensity
and the line center and the line width increase with rotational
velocity. 
When the rotational velocity is high enough, the line becomes single
peaked and symmetric, a unique feature that other theoretical models
find difficult to reproduce without introducing more complexity into
the gas distribution.

@@

\tol\ is a compact star forming dwarf galaxy that presents a
strong Ly-$\alpha$ emission \cite{Thuan97} with two puzzling 
features: it is symmetric and single peaked.
Commonly, the Ly-$\alpha$ line has a single or asymmetric double peak. 
These two special features in \tol\ had evaded a physical
interpretation so far. Using either bulk rotation or a multiphase ISM,
we manage to explain the \lya\ line in \tol.

Figure 1. summarizes our findings.
Dots represent the observational data for \tol\ with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continuous line) and a clumpy ISM
(dashed line).

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{-60}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $M_{\rm
  dyn}=5.9^{+3.1}_{-3.4}\times 10^{9}$ $M_{\odot}$.

In the rotation model assuming reasonable bounds for the number
density of neutral Hydrogen atoms (
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$) and using $\tau=10^7$ the radius
of the emission region can be bracketed to be in $0.55 <
r_s/\mathrm{kpc}< 5.5$.
Using $V_{\rm max}=300$km s$^{-1}$, we find a dynamical mass range of
$5.45\times 10^{9}M_{\odot} < M_{\rm dyn} < 5.45\times 10^{10} M_{\odot}$.

Radio surveys of the 21 cm line have put an upper limit to the neutral
hydrogen mass in \tol\ of $M<2.65\times 10^{8}$ M$_{\odot}$
\cite{pustilnikmartin07}.  
Our models thus provide evidence for a dynamical mass at least 20 times
higher than the mass in neutral hydrogen.   
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars has been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.  
Ionization emission lines present in this galaxy support the idea that
\tol\ might have harbored an an Active Galactic Nucleus.

@@

From the multiphase model we obtained $v=71_{-25}^{+17}$ \kms
and $r=5$ kpc. 
This corresponds to a dynamical mass of $M_{\rm
  dyn}=5.9^{+3.1}_{-3.4}\times 10^{9}$ $M_{\odot}$ which is at least
$20$ times the HI mass estimated from observations.

In the rotation model the size of the spherical region, $r_{s}$, can
be infered from the relationship $\tau = \sigma_0 n r_s$, where
$\sigma_0=5.898\times 10^{-14}$cm$^{-2}$ is the Lyman$\alpha$
cross section at the  line's center, $n$ is the number density. 
This expresion can be rewritten as
\begin{equation}
r_{s} = 0.055 \left(\frac{\tau}{10^7}\right) \left(
\frac{\mathrm{atoms\ cm^{-3}}}{n}\right) \mathrm{kpc}.
\end{equation}
%
Using this result and assuming a bound for the number density of
$0.01<n/\mathrm{atoms\ cm^{-3}}<0.1$ we have $0.55
< r_s/\mathrm{kpc}<5.5$.

@@

Using those size bounds for the spherical model we find a constrain for
the dynamical mass of $5.45\times 10^{9}M_{\odot} < M_{\rm dyn} <
5.45\times 10^{10} M_{\odot}$.
